### 2005-12-20 inst/paper/paper2.tex
Added new pictures and the bibliography file and updated the paper.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@15482 bc3139a8-67e5-0310-9ffc-ced21a209358

\begin{titlepage}
\begin{center}
\Large
\vspace*{1in}
\textbf{Creating Linked, Interactive Views to \linebreak Explore Linked Data} \\
\vspace{1in}
\textbf{Dana-Farber Cancer Institute}\\ 
\textbf{December 9, 2005} \\
\vspace{1in} 
\textbf{Elizabeth Whalen} \\
\end{center}
\end{titlepage}
\normalsize
\newpage

@@

There are many ways to visualize multivariate data where more than
two or three dimensions are needed to view the data.  Several of these ideas
were discussed in a previous paper, titled ``Creating Linked, Interactive
Views to Explore Multivariate Data'' \cite{EW05}, so the focus in this paper
will be one specific method of visualizing multivariate data that is
particularly useful for exploratory data analysis.  This paper will discuss
creating linked, interactive views of linked data sets.

First brief definitions of linked views, interactive views and linked data
sets will be given.  Linked views mean that if a component that represents
data (such as a point on a plot or a row in a spreadsheet) changes its
appearance on one view, then the corresponding component on a second view will
also change its appearance.  For components on different views to be
corresponding, the data displayed in those components must come from the same
entity in a model.  As an example, suppose that the model consisted of four
variables for the fifty states' murder rate, assault rate, rape rate, and
percent of population living in urban areas.  Now the model could be
represented as a matrix of four columns by fifty rows with four columns for
the four variables and fifty rows for the fifty states.  In this model, the
entities would be the states.  If two scatterplots were created as views of
this model, where the first plot displayed murder rate versus assault rate and
the second plot displayed rape rate versus percent urban population, then the
points on each plot that referred to the state of Montana would be
corresponding components.

Interactivity is an idea that will be familiar from using the Internet.  For
example, clicking on a hyperlink on a web page results in a new web page
opening.  Another example is clicking on the login button after entering a
user name and password to view one's email.  Interactivity means that there is
a response to some user action.

One particular model for creating interactive applications involves three
components: some user action or input, which will be referred to as an event;
a response to that action, which will be executed by a function that is called
a callback function; and a method that connects the event to the callback
function, which will be referred to as the signal handler.  The events can
range from the user moving the cursor over a plot to pressing a button on a
graphical user interface (GUI) to pressing certain keys.  Notice that the
events almost always occur over some type of interface, though some events,
such as key press events may occur without an interface.  The callback
function is a unique function because its signature is determined by the event
that the callback function is linked to.  The parameters of the callback
function may include information about the event, such as the location where
the event occurred, and information about the object that just caused the
event, such as the button on the GUI.  The signal handler is the piece that
must look to see if an event has occurred and then when the event has
occurred, the signal handler calls the appropriate callback function.  The
strengths of this model of interactivity are that programmers can decide which
events to notice by deciding which signal handlers to define, programmers
can change the response to an event by changing the callback function that is
linked to an event, and the system's event loop can continue working on other
things until a signal is emitted because an event occurred.

These three pieces in this model of interactivity are ordered as follows:
an event occurs, the event causes a signal to be emitted that is caught by the
signal handler, and then the signal handler calls the callback function, which
results in the response to the user action.  A flowchart of these steps is
shown below in Figure \ref{Fig:Event}.  In the example of clicking on a
hyperlink mentioned above, the event is the user clicking on the
hyperlink and the callback function causes a new web page to open.
The signal handler, which is the method that connects the event to the
callback function, is not noticed by the user, but it is handled by the
system's event loop when a signal is emitted.

\clearpage

@@

Linked data sets mean that the data have more than one conceptual grouping,
but there is a relationship between these different conceptual groupings.
Linked data sets are an idea that will be familiar to users of relational
databases.  Data in a relational database are typically stored in many
linked tables with rows in different tables linked to each other through a
unique identifier called a key.  In this example, each table would be
considered a data set and they are linked through the keys.  Another situation
where linked data sets occur is when there is experimental or study data that
are linked to meta-data.  An example of this situation is given in Section
\ref{Ssec:Limit} where the experimental data are microarray data, which looks
at gene expression levels, and the meta-data are a Gene Ontology graph that
gives the molecular functions of the genes studied in the microarray
experiment.

Creating linked, interactive views of data is flexible for exploring
multivariate data because many different views can be created, such as
scatterplots, spreadsheets, and heatmaps, and these views are linked because
they are based on the same underlying model.  Thus, users can decide which
views best visually represent their data.  Also, since the views are
interactive, users can change the views while looking at them to make the
views more informative about the underlying data structure.  This flexibility
in linked, interactive views gives users a powerful tool for visually
exploring their data.

In Section \ref{Ssec:Design}, the design of linked, interactive views based on
the model-view-controller (MVC) paradigm will be discussed.  However, the
limitation with this design is that it expects one data set, which limits the
type of data we can visualize.  There are many
situations where multiple data sets occur, particularly any time there are
experimental or study data linked to meta-data.  Being able to visualize data
from multiple data sets would be helpful to understand the relationships
between variables.  Thus, after discussing the limitations of the MVC design
in Section \ref{Ssec:Limit}, an extension of the MVC design will be discussed
in Section \ref{Sec:Extend}.  This extension will allow the creation of linked,
interactive views based on linked data sets.  The software that is discussed
in this paper will have three goals: to create linked views of linked data
sets, to create interactive views where the response to an event can be
changed, and to create an extensible design that will allow future users to
make additions.

As a motivating example of linked views based on linked data sets, consider
Figure \ref{Fig:firstMP}.  In this figure, there are two views, one view of
data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
When a user initially interacts with the scatterplot view of data set 2, the
response to this interaction is to tell data set 2 to update itself (this is
indicated on the figure as a circled 1).  Then in step 2, data set 2 tells the
scatterplot view that is should update itself in response to the change in the
data.  Next, in step 3, because data set 2 has changed and because it is
linked to data set 1, data set 2 sends information to data set 1 about the
change.  In step 4, the information from data set 2 is converted into
information that data set 1 can understand.  In step 5, this information is
received by data set 1 and data set 1 updates its data.  Finally, in step 6,
data set 1 tells its view to update itself.  Thus, in this example, the views
are synchronized based on the underlying linked data sets.

\clearpage

@@

@@

Focusing on linked, interactive views as a powerful visualization tool, the
next step is to determine how to design a software package that can implement
this method.  A popular design for creating linked views of data is the
model-view-controller (MVC) paradigm.  This design consists of three types
of objects: the controller, which defines what actions occur in response to
user input; the view, which consists of displays of the data; and the model,
which manages the data.  This paradigm is powerful because it decouples the
views from the model.  Thus, only one copy of the data needs to be stored and
then all views refer to this data set.  With all views based on the same data,
any changes to the data can be propagated to all views, which will create
linked views.

To link the views to this one data set there is a subscribe/notify procedure
between the views and the model.  The views subscribe to a particular model
and the model must notify the views when a change occurs so that the views
will be updated.  This design allows multiple views of the model that are
linked, but are not aware of each other, as shown in Figure \ref{Fig:ExMVC}
\cite{DesignPatterns}.  Each view depends only on visually representing the
model and thus, it is not necessary for a view to be aware of other views.  By
keeping each view ignorant of other views, other views can added and removed
without affecting a different view.

@@

Notice that although the components (model, view, and controller) in the MVC
design are independent of one other, there still needs to be communication
between them so that information can be shared.  Examples of this
communication between components include the model notifying views that the
data have changed, a view asking the controller how it should respond to user
interaction, and a view telling the model that the data should change in
response to user interaction with the view.  This message passing between
components in the MVC paradigm will be discussed in detail in Section
\ref{Ssec:OneMess}.

The ``Creating Linked, Interactive Views to Explore Multivariate Data''
\cite{EW05} paper discussed the creation of R software to fulfill three goals:
to create views that are linked, to create views that are interactive, and to
create a design that is extensible.  To satisfy the overall goals of the
software, two R packages, called \Rpackage{MVCClass} and \Rpackage{iSPlot},
were created.  The \Rpackage{MVCClass} package implemented the class and
generic function definitions needed for the MVC design and the
\Rpackage{iSPlot} package created the front end to let the user create linked,
interactive views of two-dimensional data, such as data frames and matrices,
through either a GUI or through command line functions.  The reason for
separating the software into two packages was that it allowed the class
definitions in the \Rpackage{MVCClass} package to be used by any other R
package that would like to create linked views.

@@

The MVC design assumes that there is one data set that all views will
reference.  But what if the data consisted of more than one conceptual
grouping?  For example, consider data stored in a relational database, where
there are multiple tables of data and rows in different tables are linked
through a unique identifier called a key.  Although these tables of data could
be combined into one large table, there are several reasons not to do this.
Keeping the data in separate tables reduces redundancy, particularly when the
rows are related by a one-to-many or a many-to-many relationship.  Also
separate data tables generally lead to more maintainable data.  By storing
only one copy of each piece of information (in a row in one table), it is
easier to ensure that the data are correctly updated when a change occurs.
All other tables that are connected with this information can then just
reference this row that has been updated.

@@

Clearly, this information could be combined into one large table, called
CombinedStateCity, as shown in Figure \ref{Fig:OneDBTab}.  The
CombinedStateCity table would have the following columns: State, Life
Expectancy, Income, City, Precipitation, and Temperature.  Now if a state's
life expectancy changed, all instances of this state would need to be changed
in the new large table.  For example, if the state had the information for
three cities recorded, then the state would have three rows in the
CombinedStateCity table and each row would need to change the value for the
state's life expectancy.  If instead only one record was kept per state, as
shown in Figure \ref{Fig:DBTab} above, then this process would be less error
prone and would require less memory to store the data.  This example shows
that there are advantages to storing the data in separate structures.

@@

In a relational database all of the data are stored in tables, but consider the
situation where there are multiple data sets that are of different data
structures.  For example, suppose that you have experimental microarray data,
which look at gene expression levels.  This experimental data can be
stored as a two dimensional data structure, such as a matrix, where rows are
genes and columns are samples.  Linked to this experimental data are meta-data
that give information about the genes studied in the microarray experiment.
For example, this meta-data may be a Gene Ontology graph that gives the
molecular functions of these genes.  These two data sets (the experimental
data and the meta-data) are linked through the genes.  Each gene can have more
than one molecular function and each molecular function can have more than one
gene annotated at that function so there is a many-to-many relationship
between the genes and the molecular functions.  Though these two data sets are
linked, the microarray data and the Gene Ontology data are different data
structures.  The experimental microarray data can be represented as a matrix
and the Gene Ontology meta-data can be represented as a graph.

Now for the original MVC design, where one data set is expected, it is not
clear how to combine these two data sets into one data set.  The meta-data
graph could be stored as a matrix, but a graph is a data structure that can be
well represented as an object with slots for nodes and edges, especially if
the graph is sparse.  As with the relational database tables, it seems
best in terms of data storage, alteration, and retrieval to keep the two data
sets separate.  Also, keeping the two data sets separate does not force a data
structure on the data that may not fit well.

@@

As discussed in Section \ref{Sec:Intro}, the MVC design expects a single
conceptual unit, such as a data frame or a matrix,
with all views referencing this one data set.  However, in Section
\ref{Ssec:Limit} several examples were shown where multiple data sets are
linked.  These examples included linked tables in relational databases and
experimental data linked to meta-data.  Being able to visualize variables
simultaneously from different data sets that are linked would be useful for
exploring relationships between these variables.  Thus, the next step will be
to determine how the MVC design can be revised to account for multiple data
sets.

@@

One possibility for dealing with multiple data sets within the MVC design is to
combine the multiple data sets into one large data set.  With this new
combined data, the original MVC design is appropriate for creating linked
views.  Thus, the strength of combining the data sets into one large data set
is that the MVC paradigm is already appropriately set up to deal with the
one large data set.

Unfortunately, as discussed in Section \ref{Ssec:Limit}, combining
multiple data sets into one large data set is not always appropriate or even
possible.  Even if the multiple data sets have the same data structure,
combining the data sets often results in redundant data and makes updating the
data more complicated.  In the situation where the data sets are not of the
same data structure, it may not be possible to combine the data sets into one
data structure that retains all of the original information.  Also, combining
the data into one data structure may give the appearance that some variables
have a relationship that does not exist.  For example, suppose that in a
relational database, table 1's rows are linked to table 2's rows and table 3's
rows are also linked to table 2's rows.  Just because table 1 and table 3 are
both linked to table 2 does not necessarily mean that there is a meaningful
relationship between the data in table 1 and table 3.  However, by combining
the data in these three tables, it would give the appearance that there is a
relationship between the data in table 1 and table 3.  Thus, another
way of linking views based on multiple data sets needs to be designed.

@@

Because of the difficulty of combining multiple data sets, the MVC paradigm
will need to be altered to fit the data.  It has been mentioned that the MVC
design is intended to deal with one data set.  A more straightforward approach
may be to link MVCs.  In linked MVCs, each MVC can be coherent and the
interactions between MVCs can be encapsulated in a set of messages.  One MVC
object to handle each data set.  Thus, within each MVC object, one could
create linked views of that particular data set, which is the same as the
original MVC design.  Then between MVC objects a messaging system would need
to be designed so that a change in one data set would be sent to the other
linked data sets.

@@

\normalsize
As an example, suppose that one has two data sets: one with information at the
state level and one with information at the city level.  These tables are
shown in Table \ref{Tab:CityState}.  The State data include Life Expectancy
(in years from 1970) and Income (per capita from 1974) data
while the City data include Precipitation (in inches from 1975) and Average
High Temperature in January (in Fahrenheit from 2005)
data.  These two data sets are linked by the CityState table that links cities
with the states they reside in.  Now for these two data sets there would be
two MVC objects.  One MVC object for the State data and its views and one MVC
object for the City data and its views.

@@

This solution uses the power and simplicity of the MVC design and expands it
to the multiple data set problem.  Now we can build on the original design,
which has been well tested, instead of starting a new design from scratch.
The strength of this multiple linked MVC design is that it uses MVC
components, which can be individually created and then linked.  Thus, these
components are reusable.  Also this multiple MVC design is structured to fit
with the linking seen in the multiple data sets.  Clearly, the design should
be built around the problem and multiple MVC objects with messaging between
the MVC objects fits the problem of creating linked views of multiple linked
data sets.

Two parts of this design will need some special thought: creating an
appropriate messaging system between MVC objects and deciding how the MVC
objects will be related.  In designing an appropriate messaging system between
MVC objects, consideration will be given to what information should pass
between MVC objects and to how messages will propagate between MVC objects.
In deciding the relationship between the MVC objects, consideration will be
given to how the multiple data sets are related.  Clearly, the relationship
between the MVC objects should reflect the relationship between the data
sets.  The two examples of linked data sets that have been discussed
previously were linked tables in a relational database, and experimental data
that are linked to meta-data.  In this paper, the focus will be on experimental
data that are linked to meta-data, as those are the data we are most
interested in visualizing.  In this situation, the experimental data can be
thought of as determining what meta-data are studied.  In the example of
microarray experiment data linked to Gene Ontology meta-data, only Gene
Ontology data that refer to genes in the microarray experiment will be
included in the meta-data.  Thus, there is a hierarchical relationship between
these data sets.  Therefore, the design of the linking between MVC objects
will be based on a hierarchical relationship between the data sets, where one
MVC object creates another MVC object.  These decisions will be discussed in
the following sections as the design for multiple MVC objects is explained.

@@

\subsection{Overview}\label{SSec:OneOver}

Implementing an extensible design to create linked, interactive views of
linked data sets will be done in the R language in two packages, called
\Rpackage{MVCClass} and \Rpackage{iSNetwork}.  The \Rpackage{MVCClass} package
will implement the design of the MVC classes and generic functions.  The
\Rpackage{MVCClass} package will define the individual model, view, and
controller classes as well as the \Robject{MVC} class that binds the
individual model, view, and controller classes together into one unit.  The
package will also define the message classes that will allow the components to
communicate with each other.  Thus, the \Rpackage{MVCClass} package will
contain all of the classes necessary for creating linked views.   The goal of
separating these class and generic function definitions into one package is
that they will be available for other R packages that implement linked views
of data, such as the \Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.

With the class definitions created in the \Rpackage{MVCClass} package, the
\Rpackage{iSNetwork} package can decide which classes from \Rpackage{MVCClass}
to use based on the overall goals of the combined software.  The
\Rpackage{iSNetwork} package will also determine how objects of these classes
will behave because the methods for the classes are defined in the
\Rpackage{iSNetwork} package.  In addition to implementing the behavior of
objects, the \Rpackage{iSNetwork} package will also create a graphical user
interface (GUI) so that users can load data, create views, and interact with
the views through menus on the user interface.  As an alternative to the GUI,
users will also have access to the same functionality through command line
functions.

To create the GUI, \Rpackage{iSNetwork} will use the R packages,
\Rpackage{RGtk} and \Rpackage{gtkDevice}.  \Rpackage{RGtk} is an R package
that allows one to interact with Gtk functions from the R interface.  Gtk is
an open-source X Window toolkit \cite{Gtk} for creating user interfaces.  The
\Rpackage{gtkDevice} package creates a Gtk device by turning a Gtk drawing
area into a device that acts like an R device, but still responds to events so
that the view can be interactive.  For more information about these packages,
please see the paper, ``Creating Linked, Interactive Views to Explore
Multivariate Data'' \cite{EW05}.

>>>>

\begin{titlepage}
\begin{center}
\Large
\vspace*{1in}
\textbf{Creating Linked, Interactive Views to \linebreak Explore Linked Data} \\
\vspace{1in}
\textbf{Dana-Farber Cancer Institute}\\ 
\textbf{December 19, 2005} \\
\vspace{1in} 
\textbf{Elizabeth Whalen} \\
\end{center}
\end{titlepage}
\normalsize
\newpage

@@

There are many ways to visualize multivariate data where more than
two or three dimensions are needed to view the data.  Several of these ideas
were discussed in a previous paper, titled ``Creating Linked, Interactive
Views to Explore Multivariate Data'' \cite{EW05}, so the focus in this paper
is on one specific method of visualizing multivariate data that is
particularly useful for exploratory data analysis; creating linked,
interactive views of linked data sets.

First brief definitions of linked views, interactive views and linked data
sets are given.  Linked views mean that if a component that represents
data (such as a point on a plot or a row in a spreadsheet) changes its
appearance on one view, then the corresponding component on a second view
also changes its appearance.  For components on different views to be
corresponding, the data displayed in those components must come from the same
entity in a model.  As an example, suppose that the model consisted of four
variables for the fifty states' murder rate, assault rate, rape rate, and
percent of population living in urban areas.  Now the model could be
represented as a matrix of four columns by fifty rows with four columns for
the four variables and fifty rows for the fifty states.  In this model, the
entities are the states.  If two scatterplots were created as views of
this model, where the first plot displayed murder rate versus assault rate and
the second plot displayed rape rate versus percent urban population, then the
points on each plot that referred to the state of Montana are
corresponding components.   Linked views, which have a long history in
statistics, are now considered a standard feature of interactive data
visualization software \cite{GGobi}.  Linked views have been implemented in
several pieces of statistical software, such as XLisp-Stat and GGobi.
Implementing linked views is based on the model-view-controller (MVC) design
\cite{DesignPatterns}, which is a widely used and well understood paradigm.
The MVC design is discussed in more detail in Section \ref{Ssec:Design}.

Interactivity is an idea that is familiar from using the Internet.  For
example, clicking on a hyperlink on a web page results in a new web page
opening.  Another example is clicking on the login button after entering a
user name and password to view one's email.  Interactivity means that there is
a response to some user action.  It is a powerful paradigm and allows users to
quickly navigate and explore complex documents or complex data sets.  As an
example, suppose that a user has two views: one view is a graph of molecular
functions for genes and the other view is a heatmap of gene expression data.
If the user clicks on a node that represents a particular molecular function,
then the response to this action could be to highlit all genes in the heatmap
view that express this molecular function.  This interactivity allows the
user to quickly see if genes with the same molecular function have similar
expression levels.

One particular model for creating interactive applications involves three
components: some user action or input, which is referred to as an event;
a response to that action, which is executed by a function that is called
a callback function; and a method that connects the event to the callback
function, which is referred to as the signal handler.  The events can
range from the user moving the cursor over a plot, to pressing a button on a
graphical user interface (GUI), to pressing certain keys.  Events are
interpreted based on which window has focus.  For example, in the XWindows
system, if a user types C on the keyboard and the window that has focus is a
text editor, then the letter C is printed on the text editor's screen.
Any other open windows that do not have
focus do not respond to this event.  The callback function is a unique
function because its signature is determined by the event that the callback
function is linked to.  The parameters of the callback function may include
information about the event, such as the on screen location of the event, and
information about the object that had focus when the event occurred, such as a
button on the GUI.  The signal handler is the piece that must look to see if
an event has occurred and then when the event has occurred, the signal handler
calls the appropriate callback function.

The strengths of this model of interactivity are that programmers can decide
which events to react to by deciding which signal handlers to implement,
programmers can change the response to an event by changing the callback
function that is linked to an event, and the system's event loop can continue
working on other things until a signal is emitted when an event occurs.  A
system's event loop controls how work is processed.  The event loop checks for
work to do and handles that work.  In the model of interactivity described
above when a signal handler is implemented by the programmer, this results in
the event loop periodically checking to see if an event has occurred.  Thus,
the event loop can perform other work and then check back to see if an event
has occurred.

These three pieces in this model of interactivity are ordered as follows:
an event occurs, the event causes a signal to be emitted that is caught by the
signal handler, and then the signal handler calls the callback function, which
results in the response to the user action.  A flowchart of these steps is
shown below in Figure \ref{Fig:Event}.  In the example of clicking on a
hyperlink mentioned above, the event is the user clicking on the
hyperlink and the callback function causes a new web page to open.
The signal handler, which is the method that connects the event to the
callback function, is handled by the system's event loop when a signal is
emitted.

@@

Linked data sets mean that the data have more than one conceptual grouping,
but there is a relationship between these different conceptual groupings.
Linked data sets are an idea that is familiar to users of relational
databases.  Data in a relational database are typically stored in many
linked tables with rows in different tables linked to each other through a
unique identifier called a key.  In this example, each table is
considered a data set and they are linked through the keys.  Another situation
where linked data sets occur is when there is experimental or study data that
are linked to meta-data.  An example of this situation is given in Section
\ref{Ssec:Limit} where the experimental data are microarray data, which looks
at gene expression levels, and the meta-data are a Gene Ontology graph that
gives the molecular functions of the genes studied in the microarray
experiment.

Creating linked, interactive views of data is flexible for exploring
multivariate data because many different views can be created, such as
scatterplots, spreadsheets, and heatmaps, and these views are linked because
they are based on the same underlying data.  Thus, users can decide which
views best visually represent their data.  Also, since the views are
interactive, users can change the views while looking at them to make the
views more informative about the underlying structure in the data.  This
flexibility in linked, interactive views gives users a powerful tool for
visually exploring their data.  However, linked views can not solve all
problems.  Clearly, only a few linked views can be viewed simultaneously by
the user because only a certain number of views can be shown on a computer
screen.  Thus, there is a limit to the number of dimensions that can be
represented by linked views.

As a motivating example of linked views based on linked data sets, consider
Figure \ref{Fig:firstMP}.  In this figure, there are two views, one view of
data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
When a user initially interacts with the scatterplot view of data set 2, the
response to this interaction is to tell data set 2 to update itself (this is
indicated on the figure as a circled 1).  Then in step 2, data set 2 tells the
scatterplot view that is should update itself in response to the change in the
data.  Next, in step 3, because data set 2 has changed and because it is
linked to data set 1, data set 2 sends information to data set 1 about the
change.  In step 4, the information from data set 2 is converted into
information that data set 1 can understand.  In step 5, this information is
received by data set 1 and data set 1 updates its data.  Finally, in step 6,
data set 1 tells its view to update itself.  This example shows the components
and the communication between the components that are necessary to create
linked views of underlying linked data sets.

In the design presented in this paper, the components consist of MVC objects
and communication within and between MVC objects is performed by message
objects.  This solution uses the power and simplicity of the MVC design and
expands it to the multiple data set problem.  Now we can build on the original
MVC design, which has been well tested, instead of starting a new design from
scratch.  The strength of this multiple linked MVC design is that it uses MVC
components, which can be individually created and then linked.  Thus, these
components are reusable.  These design decisions are discussed further in
Sections \ref{Sec:OneMVC} and \ref{Sec:MultMVC}.

@@

In Section \ref{Ssec:Design}, the design of linked, interactive views based on
the MVC paradigm is discussed.  However, the limitation with this design is
that it expects one data set, which limits the type of data we can visualize.
There are many situations where multiple data sets occur, particularly any
time there are experimental or study data linked to meta-data.  Being able to
visualize data from multiple data sets is helpful to understand the
relationships between variables.  Thus, after discussing the limitations of
the MVC design in Section \ref{Ssec:Limit}, an extension of the MVC design is
discussed in Section \ref{Sec:Extend}.  This extension supports the creation of
linked MVC objects.  The design of individual MVC objects is discussed in
Section \ref{Sec:OneMVC} while the implementation of linked MVC objects is
discussed in Section \ref{Sec:MultMVC}.  The goals of the software discussed
in this paper are to create linked views of linked data sets, to create
interactive views where the response to an event can be changed, and to create
an extensible design that allows future users to make additions.

@@

Focusing on linked, interactive views as a powerful visualization tool, the
next step is to determine how to design a software package that can implement
this method.  A popular design for creating linked views of data is the
MVC paradigm.  This design consists of three types of objects: the controller,
which defines what actions occur in response to user input; the view, which
consists of displays of the data; and the model, which manages the data.  This
paradigm is powerful because it decouples the views from the model.  Thus,
only one copy of the data needs to be stored and then all views refer to this
data set.  With all views based on the same data, any changes to the data can
be propagated to all views, which creates linked views.

To link the views to this one data set there is a subscribe/notify procedure
between the views and the model.  The views subscribe to a particular model
and the model must notify the views when a change occurs so that the views
are updated.  This design allows multiple views of the model that are
linked, but are not aware of each other, as shown in Figure \ref{Fig:ExMVC}.
Each view is responsible for visually representing some aspect of the model
and thus, it is not necessary for a view to be aware of other views.  By
keeping each view ignorant of other views, views can added and removed without
affecting each other.

@@

Notice that although the components (model, view, and controller) in the MVC
design are independent of one other, there needs to be communication
between them so that information can be shared.  Examples of this
communication between components include the model notifying views that the
data have changed, a view asking the controller how it should respond to user
interaction, and a view telling the model that the data should change in
response to user interaction with the view.  This message passing between
components in the MVC paradigm is discussed in detail in Section
\ref{Ssec:OneMess}.

In \cite{EW05}, we discuss the creation of R software to fulfill three goals:
to create views that are linked, to create views that are interactive, and to
create a design that is extensible.  To satisfy the overall goals of the
design, two R packages, called \Rpackage{MVCClass} and \Rpackage{iSPlot},
were created.  The \Rpackage{MVCClass} package implements the class and
generic function definitions needed for the MVC design and the
\Rpackage{iSPlot} package creates the front end to let the user create linked,
interactive views of two-dimensional data, such as data frames and matrices,
through a GUI or through command line functions.  The reason for
separating the software into two packages was that it allowed the class
definitions in the \Rpackage{MVCClass} package to be used by any other R
package that wanted to create linked views.

@@

The MVC design assumes that there is one data set, which is called the model,
that all views reference.  But problems arise when the data consist of more
than one conceptual grouping.  For example, consider data stored in a
relational database, where there are multiple tables of data and rows in
different tables are linked through a unique identifier called a key.
Although these tables of data could be combined into one large table, there
are several reasons not to do this.  Keeping the data in separate tables
reduces redundancy, particularly when the rows are related by a one-to-many or
a many-to-many relationship.  Also separate data tables generally lead to more
maintainable data.  By storing only one copy of each piece of information (in
a row in one table), it is easier to ensure that the data are correctly
updated when a change occurs.  All other tables that are connected with this
information can then just reference this row that has been updated.

@@

Clearly, this information could be combined into one large table, called
CombinedStateCity, as shown in Figure \ref{Fig:OneDBTab}.  The
CombinedStateCity table has the following columns: State, Life
Expectancy, Income, City, Precipitation, and Temperature.  Now if a state's
life expectancy changes, all instances of this state need to change
in the new large table.  For example, if the state had the information for
three cities recorded, then the state has three rows in the
CombinedStateCity table and each row needs to change the value for the
state's life expectancy.  If instead only one record was kept per state, as
shown in Figure \ref{Fig:DBTab} above, then this process is less error
prone and requires less memory to store the data.  Another consideration in
combining data sets is that it may give the appearance of a relationship
between variables that does not exist.  For example, it may not make sense to
compare the Income data stored in the State table to the Precipitation data
stored in the City table.  Combining the data tables assumes that there is a
relationship between the variables in all of the tables.  This example shows
that there are advantages to storing the data in separate structures.

@@

In a relational database all of the data are stored in tables, but consider the
situation where there are multiple data sets that are of different data
structures.  For example, suppose that you have experimental microarray data,
which look at gene expression levels.  This experimental data can be
stored as a two dimensional data structure, such as a matrix, where rows are
genes and columns are samples.  Linked to this experimental data are meta-data
that give information about the genes studied in the microarray experiment.
For example, this meta-data may be terms from Gene Ontology, which form a
directed acyclic graph, describing the molecular functions carried out by
these genes.  These two data sets (the experimental data and the meta-data)
are linked through the genes.  Each gene can have more than one molecular
function and each molecular function can have more than one gene annotated at
that function so there is a many-to-many relationship between the genes and
the molecular functions.  Though these two data sets are linked, the
microarray data and the Gene Ontology data are different data structures.

Now for the original MVC design, where one data set is expected, it is not
clear how to combine these two data sets into one data set.  The meta-data
graph could be stored as a matrix, but a graph is a data structure that can be
well represented as an object with slots for nodes and edges, especially if
the graph is sparse.  As with relational database tables, combining the data
sets complicates data storage, alteration, and retrieval.  For example, by
combining the data sets a structure may be forced on the data that may not fit
well, especially if the original data sets are instances of different classes.
In contrast, keeping the data sets stored separately in their original
structures simplifies the processes of storing, updating, and accessing the
data.

@@

As discussed in Section \ref{Sec:Intro}, the MVC design typically uses a single
conceptual unit, such as a data frame or a matrix, for the model.  In Section
\ref{Ssec:Limit} we provided several examples where it is advantageous to link
multiple data sets.  These examples included linked tables in relational
databases and experimental data linked to meta-data.  Being able to visualize
variables simultaneously from different data sets that are linked is useful for
exploring relationships between these variables.  We next describe how the MVC
design can be revised and extended to accommodate multiple data sets.

@@

One possibility for dealing with multiple data sets within the MVC design is to
combine the multiple data sets into one large data set.  With this new
combined data, the original MVC design can be used to create linked views.

Unfortunately, as discussed in Section \ref{Ssec:Limit}, combining
multiple data sets into one large data set can be problematic.  Even if the
multiple data sets have the same data structure, combining the data sets often
results in redundant data and makes updating and representing the
data more complicated.  In the situation where the data sets do not have the
same structure, it may be necessary to define a new class to appropriately
represent the combined data.  Also, combining the data into one data structure
may give the appearance that some variables have a relationship that does not
exist.  Just because two data sets are linked does not necessarily mean that
there is a meaningful relationship between their data.  Thus, another way
of linking views based on multiple data sets needs to be designed.

@@

A more straightforward approach may be to link MVCs.  Each MVC is a coherent
object and the interactions between MVCs are encapsulated in a set of
messages.  We use one MVC object to handle each data set.  Thus, within a MVC
object, one can create linked views of that particular data set, using the
standard MVC design.  But between MVC objects we design a second messaging
system so that a change in one data set is sent to the other linked data sets.

@@

\normalsize
As an example, suppose that one has two data sets: one with information at the
state level and one with information at the city level.  These tables are
shown in Table \ref{Tab:CityState}.  The State data include Life Expectancy
(in years from 1970) and Income (per capita from 1974) data
while the City data include Precipitation (in inches from 1975) and Average
High Temperature in January (in Fahrenheit from 2005)
data.  These two data sets are linked by the StateCity table that links cities
with the states they reside in.  Now for these two data sets there are
two MVC objects.  One MVC object for the State data and its views and one MVC
object for the City data and its views.

@@

Not only does this solution use the power and simplicity of the MVC design, but
this multiple MVC design is structured to fit the linking seen in the multiple
data sets.  Clearly, the design should be built around the problem and
multiple MVC objects with messaging between the MVC objects fits the problem
of creating linked views of multiple linked data sets.  In designing an
appropriate messaging system between MVC objects, consideration is given to
what information should pass between MVC objects and to how messages propagate
between MVC objects.  As for the relationship between the MVC objects, it
should reflect the relationship between the data sets.

Although two examples of linked data sets have been discussed thus far,
this paper focuses on experimental data that are linked to meta-data, as
the data we are most interested in visualizing.  In this situation, the
experimental data can be thought of as determining what meta-data are studied.
In the example of microarray experiment data linked to Gene Ontology
meta-data, only Gene Ontology data that refer to genes in the microarray
experiment are included in the meta-data.  Thus, there is a hierarchical
relationship between these data sets and the MVC objects that contain them,
where one MVC object creates another MVC object.  The original MVC
object is the parent MVC and the new MVC object that is created from the
parent MVC is the child MVC.  These decisions are discussed in the following
sections as the design for multiple MVC objects is explained.

@@

\subsection{Implementation Details}\label{SSec:OneOver}

Implementing linked, interactive views of linked data sets is done using an
object-oriented programming design for extensibility.  The \Rpackage{MVCClass}
package implements the design of the MVC classes and generic functions.  The
\Rpackage{MVCClass} package defines the individual model, view, and
controller classes as well as the \Robject{MVC} class that binds the
individual model, view, and controller classes together into one unit.  The
package also defines the message classes that allow the components to
communicate with each other.  Thus, the \Rpackage{MVCClass} package
contains the classes necessary for creating linked views.   The goal of
placing these class and generic function definitions into one package is
that they are available for other R packages that implement linked views
of data, such as the \Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.

With the class definitions created in the \Rpackage{MVCClass} package, the
\Rpackage{iSNetwork} package can decide which classes from \Rpackage{MVCClass}
to use based on the overall goals of the combined software.  The
\Rpackage{iSNetwork} package also determines how instances of these classes
behave because the methods for the classes are defined in the
\Rpackage{iSNetwork} package.  In addition to implementing the behavior of
objects, the \Rpackage{iSNetwork} package also creates a graphical user
interface (GUI) so that users can load data, create views, and interact with
the views through menus on the user interface.  As an alternative to the GUI,
users also have access to the same functionality through command line
functions.

To create the GUI, \Rpackage{iSNetwork} uses the R packages,
\Rpackage{RGtk} and \Rpackage{gtkDevice}.  \Rpackage{RGtk} is an R package
that allows one to interact with Gtk functions from the R interface.  Gtk is
an open-source X Window toolkit \cite{Gtk} for creating user interfaces.  The
\Rpackage{gtkDevice} package creates a Gtk device by turning a Gtk drawing
area into a device that acts like an R device, but still responds to events so
that the view can be interactive.  For more information about these packages,
see \cite{EW05}.

The next subsections explain the class definitions and the methods that
pertain to a single MVC object.  Thus, the model classes,
the view classes, the controller, the \Robject{MVC} class (which binds the
model, view and controller classes together), and the message classes that
pass information between the three components in one MVC object are
discussed in the following subsections.  All of the class definitions are
implemented in the \Rpackage{MVCClass} package and all of the methods are
defined in the \Rpackage{iSNetwork} package.  Message class definitions that
pertain to communication between the MVC objects are discussed in Section
\ref{Sec:MultMVC}.
